---
title: "Lab 06 - Regular Expressions and Web Scraping"
output: 
  tufte::tufte_html:
    css: style.css
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(include  = TRUE)
```


# Learning goals

- Use a real world API to make queries and process the data.
- Use regular expressions to parse the information.
- Practice your GitHub skills.

# Lab description

In this lab, we will be working with the [NCBI API](https://www.ncbi.nlm.nih.gov/home/develop/api/)
to make queries and extract information using XML and regular expressions. For this lab, we will
be using the `httr`, `xml2`, and `stringr` R packages.

This markdown document should be rendered using `github_document` document ONLY and pushed to your _JSC370-labs_ repository in `lab06/README.md`.

## Question 1: How many sars-cov-2 papers?

Build an automatic counter of sars-cov-2 papers using PubMed. You will need to apply XPath as we did during the lecture to extract the number of results returned by PubMed in the following web address:

```
https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2
```

Complete the lines of code:

```{r counter-pubmed}
# Downloading the website
pub <- 'https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2'
website <- xml2::read_html(pub)
# Finding the counts
counts <- xml2::xml_find_first(website,
                               "/html/body/main/div[9]/div[2]/div[2]/div[1]/div[1]/span")
# Turning it into text
counts <- as.character(counts)
# Extracting the data using regex
stringr::str_extract(counts, "[0-9,]+")
```

 - How many sars-cov-2 papers are there?

There are 192, 677 papers.

Don't forget to commit your work!

## Question 2: Academic publications on COVID19 and Hawaii

Use the function `httr::GET()` to make the following query:

1. Baseline URL: https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi

2. Query parameters:

    - db: pubmed
    - term: covid19 hawaii
    - retmax: 1000

The parameters passed to the query are documented [here](https://www.ncbi.nlm.nih.gov/books/NBK25499/).

```{r papers-covid-hawaii}
base_url <- 'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi'

library(httr)
query_ids <- GET(
  url   = base_url,
  query = list(
    db = 'pubmed',
    term = 'covid19 hawaii',
    retmax = 1000
  )
)
# Extracting the content of the response of GET
ids <- httr::content(query_ids)
```

The query will return an XML object, we can turn it into a character list to
analyze the text directly with `as.character()`. Another way of processing the
data could be using lists with the function `xml2::as_list()`. We will skip the latter for now.

Take a look at the data, and continue with the next question (don't forget to 
commit and push your results to your GitHub repo!).

## Question 3: Get details about the articles

The Ids are wrapped around text in the following way: `<Id>... id number ...</Id>`.
we can use a regular expression that extract that information. Fill out the
following lines of code:

```{r get-ids}
# Turn the result into a character vector
ids <- as.character(ids)
# Find all the ids 
ids <- stringr::str_extract_all(ids, "<Id>\\d*</Id>")[[1]]
# Remove all the leading and trailing <Id> </Id>. Make use of "|"
ids <- stringr::str_remove_all(ids, "<Id>|</Id>")
```

With the ids in hand, we can now try to get the abstracts of the papers. As
before, we will need to coerce the contents (results) to a list using:

1. Baseline url: https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi

2. Query parameters:
    - db: pubmed
    - id: A character with all the ids separated by comma, e.g., "1232131,546464,13131"
    - retmax: 1000
    - rettype: abstract
    
**Pro-tip**: If you want `GET()` to take some element literal, wrap it around `I()` (as you would do in a formula in R). For example, the text `"123,456"` is replaced with `"123%2C456"`. If you don't want that behavior, you would need to do the following `I("123,456")`.
    
```{r get-abstracts}
base_url <- 'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi'

library(httr)
publications <- GET(
  url   = base_url,
  query = list(
    db = "pubmed",
    retmax = 1000,
    rettype = 'abstract',
    id = paste(ids, collapse=',')
    )
)

# status_code(publications)

# Turning the output into character vector
publications <- httr::content(publications)
publications_txt <- as.character(publications)
```

With this in hand, we can now analyze the data. This is also a good time for committing and pushing your work!

## Question 4: Distribution of universities, schools, and departments

Using the function `stringr::str_extract_all()` applied on `publications_txt`, capture all the terms of the form:

1.    University of ...
2.    ... Institute of ...

Write a regular expression that captures all such instances

```{r univ-institute-regex}
institution <- stringr::str_extract_all(
  publications_txt,
  "University\\s*of\\s*[[:alpha:]\\s]+|[[:alpha:]\\s]+\\s*Institute\\s*of [[:alpha:]\\s]+"
  ) 
institution <- unlist(institution)
head(as.data.frame(table(institution)))
```

Repeat the exercise and this time focus on schools and departments in the form of

1.    School of ...
2.    Department of ...

And tabulate the results

```{r school-department}
schools_and_deps <- stringr::str_extract_all(
  publications_txt,
  "School\\s*of\\s*[[:alpha:]\\s]+|Department\\s*of\\s*[[:alpha:]\\s]+"
  )
head(as.data.frame(table(schools_and_deps)))
```

## Question 5: Form a database

We want to build a dataset which includes the title and the abstract of the
paper. The title of all records is enclosed by the HTML tag `ArticleTitle`, and
the abstract by `Abstract`. 

Before applying the functions to extract text directly, it will help to process
the XML a bit. We will use the `xml2::xml_children()` function to keep one element
per id. This way, if a paper is missing the abstract, or something else, we will be able to properly match PUBMED IDS with their corresponding records.


```{r one-string-per-response}
pub_char_list <- xml2::xml_children(publications)
pub_char_list <- sapply(pub_char_list, as.character)
```

Now, extract the abstract and article title for each one of the elements of
`pub_char_list`. You can either use `sapply()` as we just did, or simply
take advantage of vectorization of `stringr::str_extract`

```{r extracting-last-bit}
abstracts <- stringr::str_extract(pub_char_list, "<Abstract>(\\n|.)*</Abstract>")
# abstracts <- stringr::str_remove_all(abstracts, "<Abstract>|</Abstract>")
abstracts <- stringr::str_remove_all(abstracts, '<(/)?[A-Za-z]+\\s?[a-zA-Z0-9=\\"\\s]+>')
abstracts <- stringr::str_replace_all(abstracts, "\\s+", ' ')
table(is.na(abstracts))
```

  - How many of these don't have an abstract? 

53 of them don't have an abstract.

Now, the title

```{r process-titles}
titles <- stringr::str_extract(pub_char_list, "<ArticleTitle>(\\n|.)*</ArticleTitle>")
# titles <- stringr::str_remove_all(titles, "<ArticleTitle>|</ArticleTitle>")
titles <- stringr::str_remove_all(titles, '<(/)?[A-Za-z]+\\S?[a-zA-Z0-9=\\"\\s]+>')

titles <- stringr::str_replace_all(titles, "\\s+", ' ')
table(is.na(titles))
```

- How many of these don't have a title ? 

None. All of them have a title.

Finally, put everything together into a single `data.frame` and use
`knitr::kable` to print the results

```{r build-db}
database <- data.frame(
  titles, abstracts
)
head(knitr::kable(database))
```

Done! Knit the document, commit, and push.

## Final Pro Tip (optional)

You can still share the HTML document on github. You can include a link in your `README.md` file as the following:

```md
View [here](https://cdn.jsdelivr.net/gh/:user/:repo@:tag/:file) 
```

For example, if we wanted to add a direct link the HTML page of lecture 6, we could do something like the following:

```md
View Week 6 Lecture [here](https://cdn.jsdelivr.net/gh/JSC370/jsc370-2023/slides/JSC370-slides-06.pdf)
```